Nuclear Waste Management and Disposal

### 2017-03-15 paper.tex
end-of-term update

\date{\today}%
\begin{abstract}
    As nuclears around the world continuously generate highly radioactive waste,
    how we manage nuclear waste is becoming more important. In this work,
    we examine a few scientific and technical issues related to nuclear waste management and disposal.
    The principles of radioactivity and nuclear fission help us to understand the sources and properties
    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
    nuclear waste management and disposal after
    the termination of the Yucca Mountain project.\\
    Word count: 7449\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Nevertheless, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as Transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:waste:classification}). In particular, WIPP does not accept spent nuclear
fuel.}

@@

Evidently, nuclear waste management and disposal is becoming a increasingly urgent matter globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates

In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by building up some
language and knowledge necessary to understand how nuclear waste is generated (Section~\ref{sec:phys}),
then apply those principles to study
the physical properties of nuclear waste (Section~\ref{sec:waste}), its management and temporary storage
(Section~\ref{sec:temp}), and the debates around its eventual disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
    Here we detail the nuclear physics underlying most of the discussions
    in this work. We will apply these results in our discussion of nuclear wastes.
    \subsection{Radioactivity\footnote{
    The derivations and in this section follows chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
    }}
    Radioactive decay is the process in which most of the known detrimental radiations are generated.
    Here we review the physical principles and mathematical formalisms behind it.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of quantum mechanics.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expectation value of) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        We can easily solve the differential equation given by Eq.~\ref{eq:lambda} to get
        the the number of nuclei as a function of time $t$
        \begin{equation}
            \label{eq:decay}
            N(t) = N_0 e^{-\lambda t},
        \end{equation}
        where $N_0$ gives the number of nuclei in the initial state at $t=0$.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 10^{8}$ Bq/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity and hal-life of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Neptunium-237 & $2.1\times 10^6$ & $6.9\times 10^{-4}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiations in nuclear waste: 
        Alpha decay, which generates alpha particles (helium-4 nuclei), Beta decay which generates beta
        particles (electrons) and gamma decay which generates gamma rays.
        The type of radiation (particle beam) generated
        have different properties and deserve separate trearments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a helium-4 atom ${}^4_2$He. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it does not change the parity of the nucleus and has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha.
        \end{equation}

        \paragraph{Beta decay}
        There are a variety of beta decay processes. Here we only describe the negative beta decay that is
        relevant in this work. The basic decay process usually comes from the decay of a neutron
        \begin{equation}
            n \rightarrow p + e^- + \bar{\nu}_e,
        \end{equation}
        where $n$ denote  a neutron, $p$ a proton, $e^-$ an electron and $\bar{\nu}$ an electron antineutrino.
        This can also happen inside a nucleus (bound neutron), represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^A_{Z+1} X'_{N+1} + e^- + \bar{\nu}_e.
        \end{equation}

@@

        \subsubsection{Daughter nuclei and subsequent decay\footnote{The derivation in this section follows that of
        \cite[chapt. 1.5]{l01}}}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are oftern radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by $e^{\lambda_B t}$ and substitute in $N_A=N_0 e^{-\lambda_A} t$ to get
        \begin{align*}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align*}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nuclei has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiations than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the activity of the sample
        may increase.

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiations with living tissues}
        \label{sec:interactions}
        Radiations causes damage to living tissues by depositing energy into the biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanism and effects are different.
        \paragraph{Alpha particles} An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons, in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy into the matter it
        interacts with. In tissues, $1$-MeV alpha particles travel through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.
        \paragraph{Beta radiation (elections)} Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremstrahlung radiations as they deccelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.
        \paragraph{Photons} Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on th energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. The intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily,
        although they do not deposit as much energy, they generally are more dangerous than alpha particles due to their
        penetration abilities.

        The energy deposition described above is an important cause of irreversible damage to sensitive molecules
        like the Deoxyribonucleic (DNA) molecules if the energ transferred is large enough.
        In the cases where the energy transferred is not as large, the radiation ionizes simpler molecules like water
        to form chemically active free radicals. The free radicals have unpaired electrons. Hence they go on to react and distrupt
        other molecules that may be crucial to the biological system.

        \subsubsection{Quantifying radiation on living tissues}
        The biological impact of radiation depends on the type, power and energy of the radiation, the
        exposure time, and the part of the body exposed. Various quantities have been introduced to
        the biological impact of radiation. Here we review three
        of these quantities: the absorbed dose, the equivalent dose and the effective dose.
        The SI units for dose is siever (Sv), which is equivalent to J/Kg, the amount of energy deposited into
        per unit mass of living tissue. Some work uses the unit gray (Gy) for abosrbed dose in medical context,
        but it is the same as sievert.

@@

        For an external $\gamma$-ray source emitting $\mathcal{A}$ million photons per second 
        of energy $E_{\gamma}$
        (in MeV), one can approximate the dose rate (dose received per unit of time) by \cite[chapt. 7]{l01}
        \begin{equation}
            \frac{dD}{dt}(\mu\text{Sv h}^{-1}) \approx \frac{\mathcal{A}\text{(MBq)}\times E_{\gamma}\text{(MeV)}}
            {6\times [r\text{(m)}]^2}.
        \end{equation}
        The approximate linear dependence on $E_{\gamma}$ is valid from less than $0.1$ MeV to several
        MeV and therefore should be good for all sources but those generated in an accelerator.\cite{my68}
        The inverse-squared dependence on distance is hardly surprising.

@@

        \subsubsection{Effects of radiation on humans}
        Short-term effects of radiation are usually caused by extensive cell death or cell damage, resulting
        from large-scale damages of critical molecules.\cite{u16}
        These effects are deterministic with a threshold,
        meaning that below a certain threshold the effect will not occur and
        the severity of the effect depends on the effective dose. Roughly, below the certain threshold,
        not enough free radicals are created and hence the repairing capabilities of cells can undo the damages.
        If the radiation level is high, the radiation may generate enough free radicals to disrupt the cells or directly
        ionize critical molecules. Some examples are gastrointestinal damage and impairment
        of fertility.\cite{u16}

        Delayed health effects occur a long time after the exposure and are usually stochastic. There is
        no apparent threshold and the severity is usually independent of the dose but
        the chance of occurrence depends on the dose.\cite{u16}
        These effects mainly result from the minor disruptions of DNA molecules since the disruptions are replicated
        as cells reproduce.
        For example, according to a review done by the UN Scientific Committee on the Effects of Atomic Radiation
        (UNSCEAR), the additional chance of dying of cancer due to radiation exposure above $100$ mSv is
        about 3 to 5 percent per sievert.\cite{unscear16}

        Hereditary effects of radiation can occur if the reproduction cells were damaged by radiation.
        The mechanism is similar to that of delayed health effects, since DNA molecules are passed on
        to offpsrings.
        UNSCEAR estimates that risk for severe hereditary effect is about $0.3-0.5$ percent per siever to
        the first generation following the radiation exposure.\cite{u16, unscear16}
\section{Nuclear waste: classification and properties}
\label{sec:waste}
    In this section, we will first look at a classification scheme of nuclear waste and then
    examine in details the source, composition and properties of the most detrimental type, high-level waste.
    \subsection{Classification of nuclear waste}
    \label{sec:waste:classification}
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the activity
    content and the half-life of the radionuclides.
    Activity content is a measure of how much radioactivity $\mathcal{A}(t)$, as defined in Eq.~\ref{eq:act},
    a given amount of waste contains.
    A form of waste may possesses a high activity content due to
    a high concentration of radionuclides and/or the presence of radionuclides with high activity.\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods. The NRC classification scheme is as follows.\cite{nrc09}

    \paragraph{High-level waste (HLW) and spent fuel}
    HLW has high activity contents and usually generates a large amount of
    heat as it radiates. The typical source of HLW is spent fuel from the nuclear reactor.
    In fact, since there is no commercial reprocessing facilities in the US (see section~\ref{sec:reproc}),
    almost all of the HLW is in the form of spent nuclear fuel from the reactor.
    \paragraph{Transuranic waste (TRU)}
    TRU is defined as materials that do not belong to high-level waste and are
    contaminated by alpha-emitting radionuclides with
    sufficiently long half-life ($>20$ years) of elements with atomic number $92$ (atomic number of uranium)
    or larger and concentration of activity larger than $3700$ Bq/g.\cite{j83,s01} The activity
    of TRU is usually low but it remains active for long period of time.

>>>>

\date{\today}%
\begin{abstract}
    As nuclear reactors around the world continuously generate highly radioactive waste,
    how we manage nuclear waste is becoming more pertinent. In this work,
    we examine a few scientific and technical issues related to nuclear waste management and disposal.
    The principles of radioactivity and nuclear fission help us to understand the sources and properties
    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
    nuclear waste management and disposal after
    the termination of the Yucca Mountain project.\\
    Word count: 7449\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Moreover, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:waste:classification}). In particular, WIPP does not accept spent nuclear
fuel.}

@@

Evidently, nuclear waste management and disposal are becoming increasingly urgent matters globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates

In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by introducing radiations and its biological effects (Section~\ref{sec:phys}); then
we examine the sources and the physical properties of nuclear waste (Section~\ref{sec:waste}),
its management and temporary storage (Section~\ref{sec:temp}), and the debates around its eventual disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
The principles of radioactivity and its biological effects play a central role in our
discussion of nuclear waste. Hence, we should first review and extend our understanding of 
radioactivity and radiations.
    \subsection{Radioactivity\footnote{
    The derivations and in this section follows chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
    }}
    Radioactive decay is the process in which most of the known detrimental radiations are generated.
    Here we review the physical principles and mathematical formalism behind it.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of quantum mechanics.
        Consider a nucleus going through radioactive decay.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expected) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        We can easily solve the differential equation given by Eq.~\ref{eq:lambda} to get
        the number of nuclei as a function of time $t$
        \begin{equation}
            \label{eq:decay}
            N(t) = N_0 e^{-\lambda t},
        \end{equation}
        where $N_0$ gives the number of nuclei in the initial state at $t=0$.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 0.01$ Ci/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity and half-life of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Neptunium-237 & $2.1\times 10^6$ & $6.9\times 10^{-4}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiations in nuclear waste: 
        Alpha decay, which generates alpha particles (helium-4 nuclei), beta minus decay which generates beta
        particles (electrons) and gamma decay which generates gamma rays.
        The type of radiation (particle beam) generated
        have different properties and deserve separate treatments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a helium-4 atom ${}^4_2$He. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha,
        \end{equation}
        where $Z$ denotes the atomic number (number of protons) of a nucleus $A$ the atomic mass, and $N$ the
        neutron count.

        \paragraph{Beta minus decay}
        There are a variety of beta decay processes. Here we only describe the negative beta decay that is
        relevant in this work. The basic decay process usually comes from the decay of a neutron
        \begin{equation}
            n \rightarrow p + e^- + \bar{\nu}_e,
        \end{equation}
        where $n$ denote  a neutron, $p$ a proton, $e^-$ an electron and $\bar{\nu}$ an electron antineutrino.
        This can also happen inside a nucleus (bound neutron), represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^A_{Z+1} X'_{N+1} + e^- + \bar{\nu}_e.
        \end{equation}

@@

        \subsubsection{Daughter nuclei and subsequent decay\footnote{The derivation in this section follows that of
        \cite[chapt. 1.5]{l01}}}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are often radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by $e^{\lambda_B t}$ and substitute in $N_A=N_0 e^{-\lambda_A} t$ to get
        \begin{align*}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align*}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nucleus has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiations than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the activity of the sample
        may increase.

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiations with living tissues}
        \label{sec:interactions}
        Radiations causes damage to living tissues by depositing energy into the biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanism and effects are different.
        \paragraph{Alpha particles} An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy as it
        travels. In tissues, $1$-MeV alpha particles travel through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.
        \paragraph{Beta radiation (elections)} Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremsstrahlung radiations as they decelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.
        \paragraph{Photons} Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on the energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. As a result, the intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily.
        Although they do not deposit as much energy as alpha particles, they are generally more dangerous due to their
        penetration abilities.

        The energy deposition described above is an important cause of irreversible damage to sensitive molecules
        like the Deoxyribonucleic (DNA) molecules if the energy transferred is large enough.
        In the cases where the energy transferred is not as large, the radiation ionizes simpler molecules like water
        to form chemically active free radicals. The free radicals have unpaired electrons. Hence they go on to react
        and disrupt
        other molecules that may be crucial to the biological system.

        \subsubsection{Quantifying radiation on living tissues}
        The biological impact of radiation depends on the type, power and energy of the radiation, the
        exposure time, and the part of the body exposed. Various quantities have been introduced to
        the biological impact of radiation. Here we review three
        of these quantities: the absorbed dose, the equivalent dose and the effective dose.
        The SI units for dose is sievert (Sv), which is equivalent to J/Kg, the amount of energy deposited into
        per unit mass of living tissue. Some work uses the unit gray (Gy) for absorbed dose in medical context,
        but it is the same as sievert.

@@

        For an external $\gamma$-ray source emitting $\mathcal{A}$ million photons per second 
        of energy $E_{\gamma}$
        (in MeV), one can approximate the dose rate (dose received per unit of time) by \cite[chapt. 7]{l01}
        \begin{equation}
            \frac{dD}{dt}(\mu\text{Sv h}^{-1}) \approx \frac{\mathcal{A}\text{(MBq)}\times E_{\gamma}\text{(MeV)}}
            {6\times [r\text{(m)}]^2}.
        \end{equation}
        The approximate linear dependence on $E_{\gamma}$ is valid from less than $0.1$ MeV to several
        MeV and therefore should be good for all sources we are considering.\cite{my68}
        The inverse-squared dependence comes from the isotropic radiation assumption.

@@

        \subsubsection{Effects of radiation on humans}
        Short-term effects of radiation are usually caused by extensive cell death or cell damage, resulting
        from large-scale damages of critical molecules.\cite{u16}
        These effects are deterministic with a threshold,
        meaning that below a certain threshold the effect will not occur and
        the severity of the effect depends on the effective dose. Roughly, below the certain threshold,
        not enough free radicals are created and hence the repairing capabilities of cells can undo the damages.
        If the radiation level is high, the radiation may generate enough free radicals to disrupt the cells or directly
        ionize critical molecules. Some examples of these short-term effects are gastrointestinal damage and impairment
        of fertility.\cite{u16}

        Delayed health effects occur a long time after the exposure and are usually stochastic. There is
        no apparent threshold and the severity is usually independent of the dose but
        the chance of occurrence depends on the dose.\cite{u16}
        These effects mainly result from the minor disruptions of DNA molecules from the free radicals generated
        by low doses of radiation. The disruptions are replicated
        as cells reproduce.
        For example, according to a review done by the UN Scientific Committee on the Effects of Atomic Radiation
        (UNSCEAR), the additional chance of dying of cancer due to radiation exposure above $100$ mSv is
        about 3 to 5 percent per sievert.\cite{unscear16}

        Hereditary effects of radiation can occur if the reproduction cells were damaged by radiation.
        The mechanism is similar to that of delayed health effects, since DNA molecules are passed on
        to offspring.
        UNSCEAR estimates that risk for severe hereditary effect is about $0.3-0.5$ percent per sievert to
        the first generation following the radiation exposure.\cite{u16, unscear16}
\section{Nuclear waste: classification and properties}
\label{sec:waste}
    In this section, we will first look at a classification scheme of nuclear waste and then
    examine in details the source, composition and properties of the most detrimental type, high-level waste.
    \subsection{Classification of nuclear waste}
    \label{sec:waste:classification}
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the averaged specific activity
    and the half-life of the radionuclides in the waste.\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods. The NRC classification scheme is as follows.\cite{nrc09}

    \paragraph{High-level waste (HLW) and spent fuel}
    HLW has high activity contents and usually generates a large amount of
    heat as it radiates. The typical source of HLW is spent fuel from the nuclear reactor.
    In fact, since there is no commercial reprocessing facilities in the U.S. (see section~\ref{sec:reproc}),
    almost all of the HLW is in the form of spent nuclear fuel from the reactor.
    \paragraph{Transuranic waste (TRU)}
    TRU is defined as materials that do not belong to high-level waste and are
    contaminated by alpha-emitting radionuclides with
    sufficiently long half-life ($>20$ years) of elements with atomic number $92$ (atomic number of uranium)
    or larger and concentration of activity larger than $3700$ Bq/g.\cite{j83,s01} The activity
    of TRU is usually low but it remains active for long period of time. Moreover, TRU can be easily turned
    into plutonium-239, the {\it de facto} material to fabricate atomic bombs.
